---
title: "Gov 50 Final Project"
author: "Dillon Browning"
description: "My final project"
output: 
  distill::distill_article:
    self_contained: false
---




## Introduction

According to the United States Census, approximately 60 million people, or 19.3% of the American population, live in rural areas. While America's rural counties have been successful in diversifying their economies and addressing certain socioeconomic disparities across the rural-urban divide, there are still significant disparities between urban and rural communities regarding economic, educational, and health outcomes. Rural America experienced a decrease in population, aging demographics, and decreasing life expectancy. Policy makes have long suggested innovative ways to address these disparities such as transportation subsidies like the Essential Air Service Program or Amtrak stops in rural communities. In this project, I will examine the impact of access to intercity transportation on the drug overdose rate and poverty. 

## Part 3 Data

GEOID: state + count identifier

```{r}
#Load libraries
library(tidyverse)
library(dplyr)

#Load transportation data set
#Save to "access"
#Rename variables for merging with drug overdose data set
access <- read.csv("Data/Access_to_Intercity_Air__Bus__and_Rail_Transportation_in_Rural_Areas.csv")|>
  rename(ST_ABBREV = STATEPO, COUNTYNAME = NAME)
```

```{r}
#Change RURALPOP column to number in access
access$RURALPOP <- as.double(access$RURALPOP)

#Change GEOID column to integer in access
access$GEOID <- as.integer(access$GEOID)

#Create access21 with data from 2021 only
#Take mean of multiple entries from different projections/observations
access21 <- access|>
  filter(YEAR  == 2021)|>
  group_by(GEOID)|>
  summarise(total_pop = as.integer(mean(as.integer(TOTPOP))), 
            rural_pop = as.integer(mean(RURALPOP)), 
            perc_sml_airport = mean(as.numeric(PACCAIR)),
            perc_large_airport = mean(as.numeric(PACCAIRL)),
            perc_bus = mean(as.numeric(PACCBUS)), 
            perc_rail = mean(as.numeric(PACCRAIL)), 
            perc_all = mean(as.numeric(PACCALL)),
            perc_poverty_intercity = mean(as.numeric(POV_ANY)),
            perc_l10k_intercity = mean(as.numeric(L10K_ANY)),
            perc_unemp_intercity = mean(as.numeric(UEMP_ANY)),
            perc_unemp = mean(as.numeric(UEMP)),
            perc_poverty = mean(as.numeric(POV)))
            
```


```{r}
#Load drug overdoses data set and save to drug_overdoses
#Rename column Provisional.Drug.Overdose.Deaths with easier name drug_deaths
drug_overdoses <- read.csv("Data/VSRR_Provisional_County-Level_Drug_Overdose_Death_Counts.csv") |>
  rename(drug_deaths = Provisional.Drug.Overdose.Deaths)

#Rename FIPS to GEOID for merging data sets
drug_overdoses <- drug_overdoses |>
  rename(GEOID = FIPS)

#Change drug_deaths column to number
drug_overdoses$drug_deaths <- as.double(drug_overdoses$drug_deaths)

#Create new column drug_overdoses data set
#drug_death 2 combines drug_death values & assumes missing values
#In places with 1-9 overdose deaths, assume 1 death
drug_overdoses <- drug_overdoses|>
  mutate(drug_death2 = if_else(Footnote == "One or more data cells have counts between 1-9 and have been suppressed in accordance with NCHS confidentiality standards.", 1, drug_deaths))
  
#Create overdoses_21 with sum of overdose deaths in 2021 by county
overdoses_2021 <- drug_overdoses |>
  filter(Year == 2021)|>  group_by(GEOID, COUNTYNAME, ST_ABBREV)|>
  summarise(deaths_2021 = sum(drug_death2))
```
```{r}
#Join access_21 & overdoses_2021 data sets by GEOID
#Select columns to be included in new data set combined
combined <- access21|>
  left_join(overdoses_2021, by = "GEOID")|>
  select(total_pop:perc_poverty, GEOID:deaths_2021)
```

```{r}
#Create new column with overdose rate per 100000 residents
#Create new column with percent of population rural 
combined <- combined |>
  mutate(overdose_rate = (deaths_2021/total_pop)*100000, perc_pop_rural = (rural_pop/total_pop)*100)
```

```{r}

combined|>
  filter(perc_all <= 40)|>
  ggplot(mapping = aes(x = perc_all, y = overdose_rate)) + geom_point() + geom_smooth(method = "loess", se = FALSE)

combined_long <- combined|>
  pivot_longer(cols = c(perc_poverty_intercity, perc_poverty), names_to = "Poverty", values_to = "value")

od <- ggplot(data = combined, mapping = aes(x = perc_pop_rural, y = overdose_rate)) + geom_point(alpha = 0.25) + geom_smooth(method = "loess", se = FALSE) + labs(x = "Percent of Rural Population", y = "Overdose Rate", title = "Percentage of a County Rural and Overdose Rate")

ggplot(data = combined, mapping = aes(x = perc_all, y = overdose_rate))+ geom_point() + geom_smooth(method = "loess", se = FALSE) + scale_x_continuous(limits = c(0, 100))

transport_poverty <- ggplot(data = combined_long, mapping = aes(x = perc_all, y = value, color =Poverty))+ geom_point(alpha = 0.25) + geom_smooth(method = "loess", se = FALSE) + scale_x_continuous(limits = c(0, 100)) + scale_y_continuous(limits = c(0, 50)) + labs(title = "Rural Poverty in Relation to Access to Intercity Transit in Rural Areas", x = "Percent of Rural Population with Access to Intercity Transit", y = "Percent of Rural Population in Poverty") + scale_color_discrete(labels = c("Percent of Rural Area in Poverty", "Percent of Rural Area in Poverty with Intercity Transit"))




Poverty_drugs <- ggplot(combined_long, aes(x = value, color = Poverty, y = overdose_rate))+ geom_point(alpha = 0.1) + geom_smooth(method = "loess", se = FALSE)  + scale_x_continuous(limits = c(0, 50)) + labs(title = "Drug Overdose Rate in Relation to Access to Intercity Transit in Rural Areas", x = "Percent of Rural Residents in Poverty in a County", y = "Drug Overdose Rate (per 100000 People)") + scale_color_discrete(labels = c("Percent of Rural Area in Poverty", "Percent of Rural Area in Poverty with Intercity Transit"))

fit <- lm(perc_all ~ overdose_rate + perc_poverty, data = combined)

transport_poverty
Poverty_drugs
od
fit
```



The overdose rate appears to be higher in urban locations in 2021, and access to intercity transport actually appears to raise the overdose rate in rural counties. My data is somewhat flawed, however, as federal privacy acts prohibit the disclosure of the exact amount of deaths in a county from overdoses in a month if between 1 and 9 deaths occurred. For these values, I assumed there was one death (as it was the minimum) which I believe effects counties with smaller populations actually under reporting the number of deaths and leading to an inaccurate OD rate for these counties. Poverty and access to intercity transport seemed to be interesting to observe, however. In counties where the rural population had between 0 and 5 percent unemployment, there was a lower OD rate in counties with access to intercity transit.
